


Lets build a model to see if there is more that can be explained the number of pnealty goal attempts.

```{r}

major <- data %>% 
 filter(competition %in% c("six_nations", "the_rugby_championship")) %>% 
  distinct(team) %>% 
  pull()

model_data <- data %>%
  mutate(team = if_else(!team %in% major, "minor_nation", team)) %>% 
  select(-competition, -date,  -lineout_throw_won_clean ,- type_of_team, - match, -penalty_goal_attempts, -try_assist, -red_card, -yellow_card, - hemisphere) %>% 
  mutate(across(where(is_character), ~ as_factor(.))) %>% 
  mutate(year= as.factor(year))  


```


```{r}
summary(model_data)
```


```{r}
alias(lm(kicks_from_hand ~ ., data = model_data))
```

There are no alias in the data so we can check for correlation







```{r}
numeric <- model_data %>% 
  select(is.numeric)
nonnumeric <- model_data %>% 
  select(kicks_from_hand, !is.numeric)
```




```{r message=FALSE, warning=FALSE}
ggpairs(nonnumeric)
```


`
team:lineout_won_own_throw

```{r}
mod5b <- lm(kicks_from_hand ~ team + penalty_goal_successful + lineout_won_own_throw + total_turnovers_gained + team:lineout_won_own_throw, data = model_data)

summary(mod5b)
```

team:total_turnovers_gained

```{r}
mod5b <- lm(kicks_from_hand ~ team + penalty_goal_successful + lineout_won_own_throw + total_turnovers_gained + team:total_turnovers_gained, data = model_data)

```


penalty_goal_successful:lineout_won_own_throw



penalty_goal_successful:total_turnovers_gained

lineout_won_own_throw:total_turnovers_gained
















```{r}
mod1a <- lm(penalty_goal_attempts ~ try_scored, data = model_data)

autoplot(mod1a)

summary(mod1a)
```

```{r}
mod1b <- lm(penalty_goal_attempts ~ kicks_from_hand, data = model_data)

autoplot(mod1b)

summary(mod1b)
```

```{r}
mod1c <- lm(penalty_goal_attempts ~ type_of_team, data = model_data)

autoplot(mod1c)

summary(mod1c)
```


mod1a 





```{r}
mod2a <- lm(penalty_goal_attempts ~ try_scored + points_scored, data = model_data)

autoplot(mod2a)

summary(mod2a)

```

```{r}
mod2b <- lm(penalty_goal_attempts ~ try_scored + kicks_from_hand,  data = model_data)

autoplot(mod2b)

summary(mod2b)
```
mod2a 


```{r}
mod3a <- lm(penalty_goal_attempts ~ try_scored + points_scored + conversion_missed, data = model_data)
```

```{r}
autoplot(mod3a)

summary(mod3a)
```

```{r}

```

