---
title: "The Team That Kicks More Wins More"
author: "Harry Williams"
date: "08/01/2021"
output: 
  html_document: 
    theme: journal
    toc: yes
    toc_float: yes
    number_sections: yes
    df_print: kable
 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

During the Autumn Nations Cup in November 2020, Eddie Jones was questioned on his team's decision to kick the ball more in matches. He answered:“The teams that kick the most win games”.^[https://www.macclesfieldrufc.co.uk/2020/12/what-should-our-game-look-like-in-2021/] This report will examine match data from the last three years and showcase the evidence and possible explanations behind the correlation between kicking frequency and win rate.

```{r Load Packages, message=FALSE, warning=FALSE, include=FALSE}
library(corrr)
library(janitor)
library(tidyverse)
library(infer)
library(janitor)
library(lubridate)
library(tsibble)
library(rvest)
library(ggfortify)
library(GGally)
library(modelr)
library(knitr)
library(kableExtra)
library(extrafont)
library(showtext)
```


## The Data 
```{r message=FALSE, warning=FALSE, include=FALSE}
# source("1_get_data/get_data.R")




international <- c(
  "australia", "new_zealand", "england", "south_africa", "ireland", "italy",
  "japan", "scotland", "wales", "france", "argentina", "georgia",
  "fiji", "romania", "usa", "tonga", "russia", "namibia",
  "samoa", "uruguay", "canada"
)


font_add_google(name = "Amatic SC", family = "amatic-sc")

# turn on showtext
showtext_auto(F)
```


```{r Load Data, message=FALSE, warning=FALSE, include=FALSE}
team_data <- read_csv("5_clean_data/team_data.csv",
  col_types =
    cols(
      season = "c"
    )
) %>%
  mutate(year = year(date), .after = season) %>%
  select(-season) %>%
  mutate(type_of_team = if_else(team %in% international, "international", "club"), .after = team) %>%
  mutate(hemisphere = case_when(
    competition %in% c("autumn_nations_cup", "european_champions_cup", "premiership", "premiership_cup", "pro_14", "six_nations", "top_14") ~ "northern",
    competition %in% c("internationals", "rugby_world_cup") ~ "mixed",
    TRUE ~ "southern"
  ), .after = competition)

number_matches <- team_data %>%
  distinct(match_id) %>%
  nrow()
```

```{r Get number of wins and kicks, message=FALSE, warning=FALSE, include=FALSE}


# Create Data showing the teams result from the match and whether they kicked more than the opposition
matches <- team_data %>%
  filter(stat == "points_scored" | stat == "kicks_from_hand") %>%
  pivot_wider(names_from = stat, values_from = values) %>%
  group_by(date, match) %>%

  # Calculate the totals for each match
  mutate(
    total_scored = sum(points_scored),
    total_kick = sum(kicks_from_hand)
  ) %>%
  ungroup() %>%

  # Subtract own point and kicks from total to get oppositions in the row
  mutate(
    opposition_points = total_scored - points_scored,
    opposition_kicks = total_kick - kicks_from_hand
  ) %>%
  select(-total_scored, -total_kick) %>%
  mutate(kick_diff = kicks_from_hand - opposition_kicks) %>%

  # Compare own stats with oppositions
  mutate(
    result = case_when(
      points_scored > opposition_points ~ "won",
      points_scored < opposition_points ~ "lost",
      points_scored == opposition_points ~ "draw"
    ),
    kicks = ifelse(kicks_from_hand > opposition_kicks, "more", "not_more")
  )
```



### Competitions

Data has been sourced from the Rugby Pass website.^[https://www.rugbypass.com/] At time of writing this includes a total of `r number_matches` professional global Rugby matches over the last 3 years. This includes both international and club rugby games from the following competitions:

```{r Create Competetion Table, echo=FALSE, message=FALSE, warning=FALSE}
team_data %>%
  select(competition, type_of_team, hemisphere) %>%
  distinct(competition, .keep_all = TRUE) %>%
  arrange(competition) %>%
  mutate_all(~ str_replace_all(., "_", " ")) %>%
  mutate_all(~ str_to_title(.)) %>%
  rename_all(~ str_replace_all(., "_", " ")) %>%
  rename_all(~ str_to_title(.)) %>%
  kbl() %>%
  kable_classic(full_width = T, html_font = "Georgia")
```

### Variables

The variables collected from each match are as follows:

```{r Creat stats table, echo=FALSE, message=FALSE, warning=FALSE}

row_1 <- team_data %>%
  distinct(stat) %>%
  arrange(stat) %>%
  mutate_all(~ str_replace_all(., "_", " ")) %>%
  mutate_all(~ str_to_title(.)) %>%
  rename_all(~ str_replace_all(., "_", " ")) %>%
  rename_all(~ str_to_title(.)) %>%
  slice(1:8)

row_2 <- team_data %>%
  distinct(stat) %>%
  arrange(stat) %>%
  mutate_all(~ str_replace_all(., "_", " ")) %>%
  mutate_all(~ str_to_title(.)) %>%
  rename_all(~ str_replace_all(., "_", " ")) %>%
  rename_all(~ str_to_title(.)) %>%
  slice(9:16)


row_3 <- team_data %>%
  distinct(stat) %>%
  arrange(stat) %>%
  mutate_all(~ str_replace_all(., "_", " ")) %>%
  mutate_all(~ str_to_title(.)) %>%
  rename_all(~ str_replace_all(., "_", " ")) %>%
  rename_all(~ str_to_title(.)) %>%
  slice(17:24)

row_4 <- team_data %>%
  distinct(stat) %>%
  arrange(stat) %>%
  mutate_all(~ str_replace_all(., "_", " ")) %>%
  mutate_all(~ str_to_title(.)) %>%
  rename_all(~ str_replace_all(., "_", " ")) %>%
  rename_all(~ str_to_title(.)) %>%
  slice(25:32)

table <- bind_cols(row_1, row_2, row_3, row_4)

names(table) <- NULL

table %>%
  kbl() %>%
  kable_classic(full_width = T, html_font = "Georgia")
```




### Three Key Variables

The first key variable explored in this investigation is `Kicks From Hand`. This is the total number of kicks in play but does not include penalty kicks. `Win Rate` is the second key variable explored, and can be defined as the percentage number of wins for a team. Finally `Kick Rate`, which is the percentage number of times a team has kicked more than their opposition. `Win Rate` and `Kick Rate` have been calculated in the following ways:

<br> 

$$Win\:Rate = \frac{Number\:of\:Matches\:Won}{Total\:Matches\:Played}$$

<br>

$$Kick\:Rate = \frac{Number\:of\:Times\:Kicked\: More\: Than\: Opposition}{Total\:Matches\:Played}$$

<br>

### Data Limitions 

#### Rugby Pass Data


```{r message=FALSE, warning=FALSE, include=FALSE}
no <- team_data %>%
  distinct(match, date, .keep_all = T) %>%
  group_by(year) %>%
  summarise(count = n()) %>%
  filter(year == 2020) %>%
  pull(count)
```

There is not an even distribution of match data available on Rugby Pass over the time period, with the majority (`r no`) taking place in 2020. Additionally, for competitions which are currently in progress, such as The Rugby Premiership and Pro 14, only the match data from the current season are included in the dataset. Previous seasons cannot be sourced. International match data are almost complete, however there are a few important matches from the World Cup (Semi-finals onwards) which are missing.

<br>

```{r echo=FALSE, message=FALSE, warning=FALSE}
team_data %>%
  distinct(match, date, .keep_all = T) %>%
  group_by(year) %>%
  summarise(Count = n()) %>%
  rename(Year = year) %>%
  kbl() %>%
  kable_classic(full_width = F, html_font = "Georgia")
```

<br>

#### Variable Limitions

`Kicks From Hand` only includes the total number of kicks made in each match. The dataset does not give a more precise count of different types of kicks in a match. Tactically, a kick from a team's own 22 is very different from a kick from the halfway line or inside the opposition's half. A breakdown of the number of kicks by territory would better reflect this. This can also be applied to penalties as it is difficult to determine how each penalty was conceded e.g. from the breakdown or scrum. This data would offer a much more detailed analysis than what the current dataset allows.



# Analysis

## Teams that kicks more than the opposition

### Percentage Number of Match Wins

Using the data, the percentage number of wins for teams that kick more than their opponents are as follows:



```{r echo=FALSE, message=FALSE, warning=FALSE}
matches %>%
  dplyr::filter(kicks == "more") %>%
  mutate(kicks = str_to_title(kicks)) %>%
  mutate(across(where(is.character), ~ str_to_title(.))) %>%
  ggplot() +
  aes(x = result, y = ..prop.., group = kicks, fill = type_of_team) +
  geom_bar(stat = "count", show.legend = FALSE) +
  facet_wrap(~type_of_team) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Results of Matches Where Teams Have Kicked More Than The Oppositon",
    x = "Result",
    y = "Percentage"
  ) +
  theme_minimal() +
  theme(text = element_text(family = "Georgia"))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
kicked_more <- matches %>%
  filter(kicks == "more")

percent_no <- kicked_more %>%
  group_by(result) %>%
  summarise(count = n(), prop = n() / nrow(kicked_more)) %>%
  filter(result == "won") %>%
  mutate(prop = paste(round(prop * 100), "%", sep = "")) %>%
  pull(prop)
```


<br> 

This graph shows the results when a team has kicked more than the opposition, split over both club and international games. The data shows that teams that kick more frequently than their opponents won around `r percent_no` of their matches.


### Hypothesis Test 

Since this dataset is a sample, a hypothesis test needs to be carried out to confirm if `r percent_no` is representative across all professional matches. The method for the test can be found in the dropdown.

<br>
<details class='maths'>
<summary>**Hypothesis Test**</summary>

If kicking had no impact on the win rate of matches, the expected proportion of matches won would be around 50%. A hypothesis test can be used to see if the dataset reflects the whole population or simply sampling variation.

The Hypothesis test can be defined as follows:

H~0~: π matches won = 0.5

H~A~: π matches won > 0.5

α = 0.05

When looking at teams that kick more than their opposition, if the p-value is greater than 0.05 then we accept H~0~. We can then conclude that the proportion of the number of matches won should be around 50%. If the p-value is less than 0.05 then H~0~ is rejected in favour of H~A~. If this is the case, we can then conclude that the proportion of the number of matches won is greater than 50%.

```{r include=FALSE}
test_matches <- matches %>%
  mutate(won = result == "won") %>%
  filter(kicks == "more")
```


```{r stats_test}

tset_matches_sample <- test_matches

null_distribution <- tset_matches_sample %>%
  specify(response = won, success = "TRUE") %>%
  hypothesize(null = "point", p = 0.5) %>%
  generate(reps = 10000, type = "simulate") %>%
  calculate(stat = "prop")

observed_stat <- tset_matches_sample %>%
  specify(response = won, success = "TRUE") %>%
  calculate(stat = "prop")
```

```{r message=FALSE, warning=FALSE}
p_value <- null_distribution %>%
  get_p_value(obs_stat = observed_stat, direction = "both")


null_distribution %>%
  visualise(bins = 30) +
  shade_p_value(obs_stat = observed_stat, direction = "right") +
  labs(title = paste("p-value =", p_value)) +
  theme(text = element_text(family = "Georgia"))
```
<br> 

The p-value is less than α, so H~0~ is rejected in favour of H~A~. There is enough evidence from the data to suggest that the proportion of wins is statistically significantly greater than the null value. This means that statistically, teams that kick more than their opponents win a larger proportion of their games.

</details>
<br>

<hr>

### Results of the test

The hypothesis test has shown that there is enough evidence from this dataset to reliably conclude that teams that kick more than their opponent will win more of their games.

A further computation can be carried out to give a percentage range within a 95% confidence interval of the population.

```{r message=FALSE, warning=FALSE, include=FALSE}
bootstrap_resample <- test_matches %>%
  specify(response = won, success = "TRUE") %>%
  generate(reps = 10000, type = "bootstrap") %>%
  calculate(stat = "prop")
```

### Estimated Percentage Wins

<br>
```{r echo=FALSE}
kick_ci_95 <- bootstrap_resample %>%
  get_confidence_interval(level = 0.95, type = "percentile") %>%
  mutate_all(~ paste(round(. * 100, 2), "%", sep = ""))


names(kick_ci_95) <- c("Lower Confidence Interval", "Upper Confidence Interval")
```

```{r include=FALSE}
lower_ci <- kick_ci_95 %>%
  pull(`Lower Confidence Interval`)

upper_ci <- kick_ci_95 %>%
  pull(`Upper Confidence Interval`)
```

```{r echo=FALSE}
kick_ci_95 %>%
  kbl() %>%
  kable_classic(full_width = F, html_font = "Georgia")
```


<br>

From this dataset, it is estimated that teams that kick more than their opponents win between `r lower_ci` - `r upper_ci` of their total matches within a 95% confidence interval. 

## Win Rate vs Kick Rate
<br>


```{r message=FALSE, warning=FALSE, include=FALSE}

# Get Win Rate
team_results <- matches %>%
  group_by(team, result) %>%
  summarise(count = n()) %>%
  ungroup() %>%
  pivot_wider(names_from = result, values_from = count) %>%
  mutate_if(is.numeric, ~ coalesce(.x, 0)) %>%
  rowwise() %>%
  mutate(
    no_matches = sum(draw, lost, won),
    win_rate = won / no_matches
  )

# Get Kick Rate
team_kicks <- matches %>%
  group_by(team, kicks) %>%
  summarise(count = n()) %>%
  ungroup() %>%
  pivot_wider(names_from = kicks, values_from = count) %>%
  mutate_if(is.numeric, ~ coalesce(.x, 0)) %>%
  rowwise() %>%
  mutate(
    no_matches = sum(more, not_more),
    kick_rate = more / no_matches
  ) %>%
  arrange(desc(kick_rate))

# Create Table
team_rates <- team_results %>%
  left_join(team_kicks) %>%
  arrange(desc(kick_rate, win_rate))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
team_rates %>%
  ggplot() +
  aes(x = kick_rate, y = win_rate) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  labs(
    title = paste("Pearson Correlation =", round(cor(team_rates$kick_rate, team_rates$win_rate), 4)),
    x = "Kick Rate",
    y = "Win Rate"
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  theme_minimal() +
  theme(text = element_text(family = "Georgia"))

cor <- round(cor(team_rates$kick_rate, team_rates$win_rate), 4)
```

<br>

Plotting `Win Rate` and `Kick Rate` there is a moderate correlation of `r cor`. While there is quite a lot of spread within the data, there is still a general trend: teams that have a higher kick rate also have a higher win rate. As there is a moderate correlation, other variables can also be used to explain `Win Rate`.

## Win Rate vs Average Number of Kicks per Game 

<br> 

### Top 10 teams by Average Number of Kicks per Game
```{r get avg no of kicks, message=FALSE, warning=FALSE, include=FALSE}
avg_kicks <- matches %>%
  group_by(team) %>%
  summarise(avg._no._kicks_per_game = mean(kicks_from_hand))

team_kicks <- team_results %>%
  left_join(avg_kicks)
```

<br> 

```{r echo=FALSE, message=FALSE, warning=FALSE}
team_kicks %>%
  filter(no_matches > 5) %>%
  ungroup() %>%
  mutate(rank = rank(-avg._no._kicks_per_game, ties.method = "first"), .before = team) %>%
  slice_max(avg._no._kicks_per_game, n = 10, with_ties = F) %>%
  select(-no_matches) %>%
  relocate(avg._no._kicks_per_game, .after = team) %>%
  mutate(team = str_replace_all(team, "_", " ")) %>%
  mutate(team = str_to_title(team)) %>%
  mutate(win_rate = paste(round(win_rate * 100, 2), "%", sep = "")) %>%
  mutate(avg._no._kicks_per_game = round(avg._no._kicks_per_game, 2)) %>%
  rename_all(~ str_replace_all(., "_", " ")) %>%
  rename_all(~ str_to_title(.)) %>%
  kbl() %>%
  kable_classic(full_width = T, html_font = "Georgia")
```

Looking at the top ten teams, there is no pattern between those with the higher average number of kicks and win rate. Four of the top ten are international teams, showing that kicking is more prevalent in the international game.

The split can be shown as follows:


<br>

```{r echo=FALSE, message=FALSE, warning=FALSE}
matches %>%
  group_by(type_of_team) %>%
  summarise(avg._no._kicks_per_games = round(mean(kicks_from_hand), 2)) %>%
  mutate(type_of_team = str_to_title(type_of_team)) %>%
  rename_all(~ str_replace_all(., "_", " ")) %>%
  rename_all(~ str_to_title(.)) %>%
  kbl() %>%
  kable_classic(full_width = F, html_font = "Georgia")
```


<br>

### Win Rate vs Average Number of Kicks per Game Plot

<br> 

```{r echo=FALSE, message=FALSE, warning=FALSE}
team_kicks %>%
  ggplot() +
  aes(x = avg._no._kicks_per_game, y = win_rate) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  labs(
    title = paste("Pearson Correlation =", round(cor(team_kicks$avg._no._kicks_per_game, team_kicks$win_rate), 4)),
    x = "Avg. No. Kicks Per Game",
    y = "Win Rate"
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  theme_minimal() +
  theme(text = element_text(family = "Georgia"))
```



<br> 

There is no correlation between `Avg. No. Kicks Per Game` and `Win Rate`. This adds a level of complexity as kicking more than the opposition can result in a higher win rate, however, simply having a high number of kicks per game does not have the same result. 

## Further Exploration 

<br>

As the data is more complete when looking at international matches, this exploration will now only focus on the international game.


```{r message=FALSE, warning=FALSE, include=FALSE}
data <- team_data %>%
  pivot_wider(names_from = stat, values_from = values) %>%
  rowwise() %>%
  mutate(penalty_goal_attempts = penalty_goal_successful + penalty_missed)
```

### Correlation with `Kicks From Hand`

Below is a graph showing which other variables correlate with `Kicks From Hand`:


```{r echo=FALSE, message=FALSE, warning=FALSE}
data %>%
  filter(type_of_team == "international") %>%
  select(-match_id:-team, -type_of_team) %>%
  correlate() %>%
  focus(kicks_from_hand) %>%
  mutate(across(where(is.character), ~ str_replace_all(., "_", " "))) %>%
  mutate(across(where(is.character), ~ str_to_title(.))) %>%
  ggplot() +
  aes(x = reorder(term, kicks_from_hand), y = kicks_from_hand, fill = kicks_from_hand > 0) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(
    title = "Other Variables vs. Kicks From Hand",
    x = "Variable",
    y = "Correlation with Kicks From Hand"
  ) +
  theme_minimal() +
  theme(text = element_text(family = "Georgia"))
```

<br> 

Any correlation score over 0.2 is weak. There appears to be a weak correlation between the number of `Kicks from Hand` and the number of `Penalty Goal Successful` and `Penalty Goal Attempts`. This could explain some of the reasons why teams that kick more than their opponent have a higher chance of winning. 

Unfortunately the data variables are not suited to make an accurate model to explain this link. However, the data can offer some insight by highlighting key variables within an explanatory model.

### Explanatory Model 

<details class='maths'>
<summary>**Model Construction**</summary>

```{r message=FALSE, warning=FALSE, include=FALSE}

major <- data %>%
  filter(competition %in% c("six_nations", "the_rugby_championship")) %>%
  distinct(team) %>%
  pull()

model_data <- data %>%
  filter(type_of_team == "international") %>%
  mutate(team = if_else(!team %in% major, "minor_nation", team)) %>%
  select(-competition, -date, -lineout_throw_won_clean, -type_of_team, -match, -penalty_goal_attempts, -try_assist, -red_card, -yellow_card, -hemisphere) %>%
  mutate(across(where(is_character), ~ as_factor(.))) %>%
  mutate(year = as.factor(year))
```



Penalty Goal Successful and team both look like good options


```{r}
mod1a <- lm(kicks_from_hand ~ team, data = model_data)
```


```{r}
autoplot(mod1a) +
  theme(text = element_text(family = "Georgia"))

summary(mod1a)
```

As with the top 10 teams with the highest average kicks per game, England, France, South Africa, Wales contribute towards kicking more in matches

```{r}
mod1b <- lm(kicks_from_hand ~ penalty_goal_successful, data = model_data)

autoplot(mod1b) +
  theme(text = element_text(family = "Georgia"))

summary(mod1b)
```

mod1a has the highest r^2 so that will be used.





```{r}
mod2a <- lm(kicks_from_hand ~ team + penalty_goal_successful, data = model_data)

autoplot(mod2a) +
  theme(text = element_text(family = "Georgia"))

summary(mod2a)
```




```{r}
mod2b <- lm(kicks_from_hand ~ team + year, data = model_data)

autoplot(mod2b) +
  theme(text = element_text(family = "Georgia"))

summary(mod2b)
```

mod2a has the highest r^2 so is the best option 



```{r}
mod3a <- lm(kicks_from_hand ~ team + penalty_goal_successful + lineout_won_own_throw, data = model_data)

autoplot(mod3a) +
  theme(text = element_text(family = "Georgia"))

summary(mod3a)
```

```{r}
mod3b <- lm(kicks_from_hand ~ team + penalty_goal_successful + total_turnovers_gained, data = model_data)

autoplot(mod3b) +
  theme(text = element_text(family = "Georgia"))

summary(mod3b)
```

mod3a  has the highest r^2 is the better model going forward.



```{r}
mod4a <- lm(kicks_from_hand ~ team + penalty_goal_successful + lineout_won_own_throw + line_break, data = model_data)

summary(mod4a)
```

```{r}
mod4b <- lm(kicks_from_hand ~ team + penalty_goal_successful + lineout_won_own_throw + total_turnovers_gained, data = model_data)

summary(mod4b)
```




```{r}
mod4c <- lm(kicks_from_hand ~ team + penalty_goal_successful + lineout_won_own_throw + lineout_won_steal, data = model_data)

autoplot(mod4c) +
  theme(text = element_text(family = "Georgia"))

summary(mod4c)
```




team:penalty_goal_successful

```{r}
mod5a <- lm(kicks_from_hand ~ team + penalty_goal_successful + lineout_won_own_throw + total_turnovers_gained + team:penalty_goal_successful, data = model_data)

summary(mod5a)
```


```{r}
anova(mod4b, mod5a)
```

mod4b is the best model as team:penalty_goal_successful is not statistically significant

```{r}
best_model <- mod4b

relaimpo::calc.relimp(best_model)
```


</details>
<br>

<hr>

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(best_model)
```

<br> 


The model summary highlights which teams are more likely to play a kicking game (the higher estimate in the summary, the more likely the team is to kick more). But, because of the spread in the data and the lack of precise variables it is not possible to draw any key conclusions. However, there is a statistically significant relationship between the number of `penalty goal attempts`, `own lineouts won` and `total turnovers gained` which suggests that there is a link between those variables.

One possible explanation could be that when a team kicks more than their opposition, the game would be played more in the opposition's territory, so therefore any penalties that are awarded would be within kicking distance. However, without looking at the type of kicks or why penalties are conceded it is difficult to come to any formal conclusions using this dataset.




# Key Takeaways
* Teams that kick more than their opposition are between `r lower_ci` - `r upper_ci` more likely to win.
* Teams that constantly kick more than their opponents are also more likely to have a higher win rate.
* There is no link between having a higher average number of kicks per game and having a higher win rate.
* Looking at international matches, the most important factor in determining the number of kicks in a match is the team. The model highlights which teams are more likely to kick the ball.
* The model also highlights possible links between the number of `kicks from hand`, `penalty goal attempts`, `own lineouts won` and `total turnovers gained`. However no formal conclusions can be made from this dataset. 



# Future Research

This dataset has led to some useful insights, but fails to explain the links between `kicks from hand`, `penalty goal attempts`, `own lineouts won` and `total turnovers gained`. A dataset that breaks down the location of kicks and the reasons for why penalties are conceded could perhaps better explain the link. Unfortunately this data cannot be scraped from the Rugby Pass website.

<br>

# References
